---
  title: "Comparison of protein vs mRNA expression in PNS development"
  output: html_notebook
---
#Outline:
Section 1 (before start of this notebook):
1. Get marker names from fcs file
2. Add to google sheet: 
  - (https://docs.google.com/spreadsheets/d/1jcHkGr2LofYwpUKI2zYrpdkZtcThkRZMQEG97klF_dk/edit#gid=0)
3. Get approved name/ensemble ids from HGNC and manual curation
4. Use ensemble ids to get corresponding gene names from biomart -> create file "pns_ab_name_translator_20200802.csv"

Section 2 (in this notebook, any pre-prepped data is included in folder on github):
5. Subset both datasets down to overlapping genes/proteins
7. Summarize (% expressing, level expression) marker expression for each of the common markers & visualize over timecourse
8. Perform in silico sorting of datasets by +/- for cannonical lineage markers & visualize expression of over markers in "sorted" populations

# Read in packages, set directories for read and write
```{r}
library(dplyr)
library(tidyr)
library(Matrix)
library(ggplot2)

READ_PATH <- getwd()
WRITE_PATH <- getwd()

#Source function files
source(paste0(READ_PATH, "PlottingFunctions_ProteinRNAComparison.R"))


```

# COMPARISON HEATMAP DOTPLOTS

### FUNCTION TO PREP SUMMARIZED DATA
```{r}
prep_for_plot <- function(exprs_path, age.metadata, markers.extract, markers.label, thresholds, file.name) {
  
  # Generate expression matrix consisting of gene levels and timepoint######################
  exp.mat <- read.csv(exprs_path)[, markers.extract]
  
  # Set colnames to be common gene names: ############################################################################
  colnames(exp.mat) <- markers.label 
  
  # Combine expression with by-cell metadata into one matrix: ################################################
  exp.mat <- cbind(exp.mat, age.metadata)
  
  # Remove extra rna timepoint: ################################################
  tmpnt_40 <- which(exp.mat$age.metadata == 40)
  if (length(tmpnt_40) >= 1) {
    exp.mat <- exp.mat[-tmpnt_40,]
  }
  
  ###normalization function ##################################################################################
  normalit<-function(x){
    (x-min(x))/(max(x)-min(x))
  }
  
  # Calculate number of cells expressing a gene for each timepoint ###########################################
  # First, read in and process "positive" cutoffs for the subset of markers that will be used for comparison
  genes_and_thresholds <- data.frame("genes" = markers.label,
                                     "thresholds" = thresholds)
  genes_and_thresholds$genes <- as.character(genes_and_thresholds$genes)

  num.cells.expressing <- apply(genes_and_thresholds, 1, function(x) {
    print(x[1])
    print(x[2])
    exprs.values <- exp.mat[,c(x[1], "age.metadata")]  %>%  
      dplyr::group_by(age.metadata) 
  
    colnames(exprs.values)[1] <- "expression"

    exprs.values <- exprs.values %>%
      dplyr::group_by(age.metadata) %>%
      dplyr::summarise(count2 = length(expression[expression > x[2]]))

    return(exprs.values$count2)
  })
  
  colnames(num.cells.expressing) <- markers.label
  
  # calculate percentage of expressing cells for color map #####################################################
  perc.expressing <- plyr::count(exp.mat, c("age.metadata")) %>% #, "class"
    dplyr::bind_cols(. ,as_tibble(num.cells.expressing)) %>%
    dplyr::mutate_each(function(v){v / .[,"freq"]}, all_of(markers.label)) %>%
    tidyr::gather(genename, percent, 3:(length(markers.label)+2), factor_key = TRUE) %>% #(would be 4 and 3 here if filtering by cell type)
    dplyr::mutate(age.metadata = as.numeric(as.character(age.metadata))) %>%
    dplyr::pull(percent)
  
  # calculate average level of expression for the genes at each timepoint #######################################
  sum.mat <- exp.mat %>%
    dplyr::group_by(age.metadata) %>%
    dplyr::select(c(age.metadata, all_of(markers.label))) %>%
    dplyr::summarise_each(funs(mean)) %>% 
    dplyr::mutate(age.metadata = as.numeric(as.character(age.metadata))) %>%
    dplyr::mutate_each(funs(normalit), all_of(markers.label)) %>%
    tidyr::gather(genename, measurement, 2:(length(markers.label)+1), factor_key = TRUE) %>%
    dplyr::mutate(percent = perc.expressing) ### merge with percent expressing cells
  
  # Add in missing timepoints with NA and/or remove excess timepoints: ################################################
  total.time.points <- c(11, 12, 13, 14, 15, 16, 17, 18, 19, 20, 21, 22, 23, 24)
  for (time.point in total.time.points) {
    if (time.point %in% unique(sum.mat$age.metadata) == FALSE) {
      for (marker.name in unique(sum.mat$genename)) {
        sum.mat[nrow(sum.mat)+1,] <- NA
        sum.mat[nrow(sum.mat),"age.metadata"] <- time.point
        sum.mat[nrow(sum.mat),"genename"] <- marker.name
      }
    }
  }
  
  sum.mat <- sum.mat[which(sum.mat$age.metadata %in% total.time.points),] #to remove p40 from rna
  
  #Write summarized expression matrix to file ##############################################################################
  sum.exprs.mean.filename <- paste0(WRITE_PATH, "0_1_range_norm_", file.name, ".csv") 
  write.csv(sum.mat, sum.exprs.mean.filename, row.names = FALSE)

}

```

### GENERATE AND SAVE SUMMARIZED DATA
```{r}

## Set read paths (but will be read in within the function):
###RNA
rna_exprs_path_in <- paste0(READ_PATH, "exprs_mat_rna_common_vars.csv")
###Protein
protein_exprs_path_in <- paste0(READ_PATH, "exprs_mat_protein_common_vars.csv")

## Read in and prep metadata files
##RNA
rna.metadata.in <- read.csv(paste0(READ_PATH, "rna_metadata.csv"))
rna.metadata.in$cell_type <- as.character(rna.metadata.in$cell_type) #make it so cell types can be str matched

rna.panel.in <- read.csv(paste0(READ_PATH, "rna_panel.csv"))
rna.panel.in$common_name <- as.character(rna.panel.in$common_name)

##PROTEIN
protein.metadata.in <- read.csv(paste0(READ_PATH, "protein_metadata.csv"))

protein.panel.in <- read.csv(paste0(READ_PATH, "protein_panel.csv"))
protein.panel.in$common_name <- as.character(protein.panel.in$common_name)
protein.panel.in$ab_name <- as.character(protein.panel.in$ab_name)
protein.panel.in$ab_metal_label <- as.character(protein.panel.in$ab_metal_label)

## Specify desired timepoints for comparison
total.time.points <- c(11, 12, 13, 14, 15, 16, 17, 18, 19, 20, 21, 22, 23, 24)

## Generate summarized files
prep_for_plot(exprs_path = rna_exprs_path_in, 
             age.metadata = as.factor(rna.metadata.in$timepoint_num), 
             markers.extract = stringr::str_to_title(rna.panel.in$common_name), 
             markers.label = rna.panel.in$common_name, 
             thresholds = rna.panel.in$thresholds, 
             file.name = "rna_summarized_age")
prep_for_plot(exprs_path = protein_exprs_path_in, 
             age.metadata = protein.metadata.in$timepoint_num, 
             markers.extract = protein.panel.in$ab_metal_label[which(protein.panel.in$summarize == 1)], 
             markers.label = protein.panel.in$common_name[which(protein.panel.in$summarize == 1)], 
             thresholds = protein.panel.in$thresholds[which(protein.panel.in$summarize == 1)], 
             file.name = "protein_summarized_age")


```

### GENERATE AND SAVE COMPARISON HEATMAP DOTPLOTS
```{r}
## Read in prepped data files
rna.sum.exprs <- read.csv(paste0(WRITE_PATH, "0_1_range_norm_", "rna_summarized_age", ".csv")) #"0_1_range_norm_", 
protein.sum.exprs <- read.csv(paste0(WRITE_PATH, "0_1_range_norm_", "protein_summarized_age", ".csv")) #"0_1_range_norm_", 

## Finesse the variable names a bit, do a confimation that they match
rna.sum.exprs$genename <- as.character(rna.sum.exprs$genename)
rna.sum.exprs$genename <- protein.panel.in$ab_name[match(rna.sum.exprs$genename, protein.panel.in$common_name)] 

protein.sum.exprs$genename <- as.character(protein.sum.exprs$genename)
protein.sum.exprs$genename <- protein.panel.in$ab_name[match(protein.sum.exprs$genename, protein.panel.in$common_name)] 

## Add column for dataset ID
rna.sum.exprs$datasetID <- as.factor(rep("R", nrow(rna.sum.exprs)))
protein.sum.exprs$datasetID <- as.factor(rep("P", nrow(protein.sum.exprs)))

## Change protein genenames to match RNA capitalization format
#protein.sum.exprs$genename <- stringr::str_to_title(protein.sum.exprs$genename)

## Bind into one df for plotting
sum.exprs.to.plot <- data.frame(rbind(rna.sum.exprs, protein.sum.exprs))
sum.exprs.to.plot <- dplyr::rename(sum.exprs.to.plot, age = age.metadata)
## Create a new column representing if there is data for a timepoint, 
#so this can be used to define point shape in ggplot
sum.exprs.to.plot$dataBinary <- rep("1", nrow(sum.exprs.to.plot))
sum.exprs.to.plot$dataBinary[which(is.na(sum.exprs.to.plot$measurement))] <- rep("0", length(which(is.na(sum.exprs.to.plot$measurement))))

sum.exprs.to.plot$measurement[which(is.na(sum.exprs.to.plot$measurement))] <- rep(0.5, length(which(is.na(sum.exprs.to.plot$measurement))))
sum.exprs.to.plot$percent[which(is.na(sum.exprs.to.plot$percent))] <- rep(1, length(which(is.na(sum.exprs.to.plot$percent))))

## Generate plot (function from "PlottingFunctions_ProteinRNAComparison.R")
gg <- plot.mean.exp.combined(sum.mat = sum.exprs.to.plot, name = "DRG Neurons")

## SAVE PLOT!
ggsave(paste0(WRITE_PATH, "figures/heatmap_dot_plots/", "comparison_tissueAgeSummarized_heatmapDotPlot_0_1_range_norm.pdf"), 
       device = 'pdf',
       plot = gg,
       width = 7.65, #6
       height = 35) #36
```



# SORTED POPULATION TIMECOURSE PLOTS
Group cells by +/- expression for key markers and plot against expression of other markers

### Read in and prep data
```{r}

#In loop, only read in the parts of the file that are necessary, then write to new sub files for faster downstream processing and testing
 
##First, set read/write paths:
###RNA
rna_exprs_path_in <- paste0(READ_PATH, "exprs_mat_rna_common_vars.csv") 
rna_subexprs_path_out <- paste0(WRITE_PATH, "exprs_subset_files/")
###Protein
protein_exprs_path_in <- paste0(READ_PATH, "exprs_mat_protein_common_vars.csv")
protein_subexprs_path_out <- paste0(WRITE_PATH, "exprs_subset_files/")


## Read in and prep metadata files
##RNA
rna.metadata.in <- read.csv(paste0(READ_PATH, "rna_metadata.csv"))
rna.metadata.in$cell_type <- as.character(rna.metadata.in$cell_type) #make it so cell types can be str matched

rna.panel.in <- read.csv(paste0(READ_PATH, "rna_panel.csv"))
rna.panel.in$common_name <- as.character(rna.panel.in$common_name)

##PROTEIN
protein.metadata.in <- read.csv(paste0(READ_PATH, "protein_metadata.csv"))

protein.panel.in <- read.csv(paste0(READ_PATH, "protein_panel.csv"))
protein.panel.in$common_name <- as.character(protein.panel.in$common_name)
protein.panel.in$ab_name <- as.character(protein.panel.in$ab_name)
protein.panel.in$ab_metal_label <- as.character(protein.panel.in$ab_metal_label)


##Now read in chunks and write to new files
###RNA
n.col.rna <- count.fields(rna_exprs_path_in, sep = ",")[1]
for (col.id in c(1:n.col.rna)) {
  rna_sub_in  = fread(rna_exprs_path_in, sep = ",", select = c(col.id) )
  col.id <- toupper(colnames(rna_sub_in)[1])
  fwrite(rna_sub_in, paste0(rna_subexprs_path_out, "pns_rna_exprs_sub_column_", col.id, ".csv"), col.names = FALSE)
  print(col.id)
}

###Protein
for (col.id in protein.panel.in$ab_metal_label) {
  protein_sub_in  = fread(protein_exprs_path_in, sep = ",", select = c(col.id) )
  col.id <- protein.panel.in$common_name[which(protein.panel.in$ab_metal_label == col.id)]
  fwrite(protein_sub_in, paste0(protein_subexprs_path_out, "pns_protein_exprs_sub_column_", col.id, ".csv"), col.names = FALSE)
  print(col.id)
}

## Get file lists for subset expression files
###RNA
rna_subexprs_path_out <- paste0(WRITE_PATH, "exprs_subset_files/")
rna.files <- list.files(rna_subexprs_path_out, full.names = TRUE)

###Protein
protein_subexprs_path_out <- paste0(WRITE_PATH, "exprs_subset_files/")
protein.files <- list.files(protein_subexprs_path_out, full.names = TRUE)


## remove CD24 and specify subset features of interest
protein.panel.in <- protein.panel.in[-c(which(protein.panel.in$ab_name == "CD24")),]
sub.list.common.names <-  c("ISL1","RBFOX3","RET","RUNX3","NTRK1","NTRK2","NTRK3","TUBB3","UCHL1") #NES",,"SOX10"
sub.list.common.names.for.sort <-  c("NTRK1","NTRK2","NTRK3")

```


### GENERATE DOT+LINE TIMECOURSE PLOTS TO SHOW SORTING
these are the first plots which show the marker we are "sorting" by, and will have all cells. subsequent violin plots will have only "positive" cells.
```{r}
##Read in expression data for all of the "sorting" markers, and combine into one df (per rna and protein)
rna.in.ls <- list()
protein.in.ls <- list()
sub.list.common.names.for.sort <-  c("NTRK1","NTRK2","NTRK3")
for (common_name in sub.list.common.names.for.sort) {
  ##RNA
  rna.in.ls[[common_name]] <- read.csv(rna.files[grep(common_name, rna.files)], header = FALSE)
  ##Protein
  protein.in.ls[[common_name]] <- read.csv(protein.files[grep(common_name, protein.files)], header = FALSE)
}
rna.toplot.df <- data.frame(rna.in.ls)
colnames(rna.toplot.df) <- names(rna.in.ls)

protein.toplot.df <- data.frame(protein.in.ls)
colnames(protein.toplot.df) <- names(protein.in.ls)

## Generate plots (function from "PlottingFunctions_ProteinRNAComparison.R")
protein_SortedLabel_LineAndDot(protein.in=protein.toplot.df,protein.metadata.in=protein.metadata.in,sub.list.common.names.for.sort=sub.list.common.names.for.sort, WRITE_PATH=WRITE_PATH)

rna_SortedLabel_LineAndDot(rna.in=rna.toplot.df,rna.metadata.in=rna.metadata.in,sub.list.common.names.for.sort=sub.list.common.names.for.sort, WRITE_PATH=WRITE_PATH)

```



### TIMECOURSE COMBINED RNA/PROTEIN LINE + VIOLIN PLOTS!
make overlayed violin and line plots of expression vs timepoint per markerX<"sorted" population>, color by type (RNA, protein)

```{r}
library(ggplot2)
library(cowplot)

# Required data was read in in previous section

names.for.plotting <- protein.panel.in$common_name[which(protein.panel.in$common_name != "CD24")]
# Now loop through and plot per sorted population
for (plot_marker in names.for.plotting) { 
  print(plot_marker)
  rna.to.plot <- read.csv(rna.files[grep(paste0(plot_marker, ".csv"), rna.files)], header = FALSE)
  protein.to.plot <- read.csv(protein.files[grep(paste0(plot_marker, ".csv"), protein.files)], header = FALSE)
  for (sort_marker in sub.list.common.names.for.sort) {
    plot_marker_name <- protein.panel.in$ab_name[which(protein.panel.in$common_name == plot_marker)]
    sort_marker_name <- protein.panel.in$ab_name[which(protein.panel.in$common_name == sort_marker)]
    plot_label <- paste0(plot_marker_name, " Expression in ", sort_marker_name, "+ Cells")
    print(plot_label)
    
    rna_violin_line_plot <- rna_SortedLabel_Violin(rna.in=rna.to.plot[which(rna.metadata.in.for.sort[,sort_marker] == 1),],
                                                              rna.metadata.in=rna.metadata.in.for.sort[which(rna.metadata.in.for.sort[,sort_marker] == 1),],
                                                              WRITE_PATH=WRITE_PATH,
                                                              plot_label = plot_label,#paste0(plot_marker, " Expression in ", sort_marker, "+ Cells"),
                                                             axis.side = "right", violin.color="blue", line.color="blue") #"skyblue"
    rna_violin <- rna_violin_line_plot[[1]]
    rna_line <- rna_violin_line_plot[[2]]


    protein_violin_line_plot <- protein_SortedLabel_Violin(rna.in=protein.to.plot[which(protein.metadata.in.for.sort[,sort_marker] == 1),],
                                                                  rna.metadata.in=protein.metadata.in.for.sort[which(protein.metadata.in.for.sort[,sort_marker] == 1),],
                                                                  WRITE_PATH=WRITE_PATH,
                                                                  plot_label = plot_label,#paste0(plot_marker, " Expression in ", sort_marker, "+ Cells"),
                                                                 axis.side = "left", violin.color="orange", line.color="orange") #"lightorange"
    protein_violin <- protein_violin_line_plot[[1]]
    protein_line <- protein_violin_line_plot[[2]]
    
    ## Align and overlay plots: #########################################################################
    aligned_plots <- align_plots(rna_violin, rna_line, protein_violin, protein_line,
                                 align="hv", axis="tblr", greedy = FALSE) #v tb h , axis="lr" tblr
    p_combined <- ggdraw(aligned_plots[[1]]) + draw_plot(aligned_plots[[2]]) + 
                  draw_plot(aligned_plots[[3]]) + draw_plot(aligned_plots[[4]])
    
    ##SAVE PLOT!
    ggsave(paste0(WRITE_PATH, "figures/violin_sorted_pop_timecourse_plots/", "combined_",  gsub(" ", "_", plot_label),  ".pdf"),  #_LINE
           device = 'pdf',
           plot = p_combined,
           width = 10,
           height = 5)
  }
}


```


